---
title: "Credit Card Record Analysis"
output: html_document
date: "2024-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Credit Card Score Analysis

Importing Libraries

```{r libraries, echo=TRUE, results=hide}
library(readxl)
library(tidyverse) # data manipulation
library(ggplot2) # grafics
library(magrittr) # %>%
library(dplyr)
library(rpart)
library(caret)
library(rsample)
library(recipes)
library(themis)     # Handling imbalanced data
```


## Importar datos

```{r, echo=FALSE}
Datos_MOD1 <- read_excel("D:\\OneDrive - Universidad San Francisco de Quito\\DATA SCIENCE\\Diplomado PUCE\\Módulo 7 - Machine Learning\\Deber 2\\default of credit card clients.xls")

```


```{r Datos, echo=TRUE, results=hide}}
View(Datos_MOD1)
```

## Data Exploration

```{r}
names(Datos_MOD1)[which(names(Datos_MOD1) == "default payment next month")] <-"default"
Datos_MOD1 %$% str(default)
```

Default refers to the expectation of the client to pay next month or not

```{r}
Datos_MOD1 %<>%
  mutate( default = factor(default,
                    levels= c("1","0"),
                    labels= c("yes", "no"))) -> Datos_MOD1
Datos_MOD1 %<>%
  mutate( SEX = factor(SEX),
          EDUCATION = factor(EDUCATION),
          MARRIAGE = factor(MARRIAGE)) -> Datos_MOD1

Datos_MOD1 %<>%
  mutate( PAY_0 = factor(PAY_0),
          PAY_2= factor(PAY_2),
          PAY_3= factor(PAY_3),
          PAY_4= factor(PAY_4),
          PAY_5= factor(PAY_5),
          PAY_6= factor(PAY_6)) -> Datos_MOD1
```


## Explore the balancing of the default payment

```{r}

Datos_MOD1 %>%
group_by(default) %>%
  summarise( Frec= n()) %>%
  mutate(Prop= Frec/ sum(Frec) ) 

```

## Summary of the variables

We include the NA values with useNA="ifany"

```{r}

summary(Datos_MOD1)

# Frequency table for the variable SEX
cat("Frequency table for the variable SEX:\n")
table_sex <- table(Datos_MOD1$SEX)
print(table_sex)

# Frequency table for the variable EDUCATION
cat("\nFrequency table for the variable EDUCATION:\n")
table_education <- table(Datos_MOD1$EDUCATION)
print(table_education)

# Frequency table for the variable MARRIAGE
cat("\nFrequency table for the variable MARRIAGE:\n")
table_marriage <- table(Datos_MOD1$MARRIAGE)
print(table_marriage)

# Frequency table for the variable PAY_0
cat("\nFrequency table for the variable PAY_0:\n")
table_pay_0 <- table(Datos_MOD1$PAY_0, useNA = "ifany")
print(table_pay_0)

# Frequency table for the variable PAY_2
cat("\nFrequency table for the variable PAY_2:\n")
table_pay_2 <- table(Datos_MOD1$PAY_2, useNA = "ifany")
print(table_pay_2)

# Frequency table for the variable PAY_3
cat("\nFrequency table for the variable PAY_3:\n")
table_pay_3 <- table(Datos_MOD1$PAY_3, useNA = "ifany")
print(table_pay_3)

# Frequency table for the variable PAY_4
cat("\nFrequency table for the variable PAY_4:\n")
table_pay_4 <- table(Datos_MOD1$PAY_4, useNA = "ifany")
print(table_pay_4)

# Frequency table for the variable PAY_5
cat("\nFrequency table for the variable PAY_5:\n")
table_pay_5 <- table(Datos_MOD1$PAY_5, useNA = "ifany")
print(table_pay_5)

# Frequency table for the variable PAY_6
cat("\nFrequency table for the variable PAY_6:\n")
table_pay_6 <- table(Datos_MOD1$PAY_6, useNA = "ifany")
print(table_pay_6)



print(paste("The mean of ages is: ",mean(Datos_MOD1$AGE)))

```

## Data Visualization

```{r}

# Create a histogram for the column "AGE" (numeric)
hist(Datos_MOD1$AGE, main = "Age Histogram", xlab = "Age")

# Create a bar plot for categorical variables (factor)
barplot(table(Datos_MOD1$SEX), main = "Gender Distribution")
barplot(table(Datos_MOD1$EDUCATION), main = "Education Distribution")
barplot(table(Datos_MOD1$MARRIAGE), main = "Marital Status Distribution")


```

##  Outliers Detection

```{r}

boxplot(Datos_MOD1$LIMIT_BAL, main = "LIMIT_BAL Boxplot")

```


## Correlations

```{r}
# Calculate the correlation matrix for numeric variables
correlation_matrix <- cor(Datos_MOD1[, c("LIMIT_BAL", "AGE", "BILL_AMT1", "BILL_AMT2", "BILL_AMT3", "BILL_AMT4", "BILL_AMT5", "BILL_AMT6", "PAY_AMT1", "PAY_AMT2", "PAY_AMT3", "PAY_AMT4", "PAY_AMT5", "PAY_AMT6")])

# Visualize the correlation matrix
heatmap(correlation_matrix)
```

## Train - Test Split

We will divide the data in 80/20 train test so we can later apply a decision tree model. The seed is set to a fix number so it can be replicated

```{r}
set.seed(1234) # Seed for randomization
data_split <- Datos_MOD1 %>%
  initial_split(prop = 0.8,
                strata = default)
train <- training(data_split)
dim(train)

test <- testing(datos_split)
dim(test)

```

## Preprocessing

```{r}
# Initialize recipe object
rct_data <- train %>% 
  recipe(default ~ ., data = ., role = "predictor") %>%
  # Normalize numeric predictors
  step_normalize(all_numeric(), -all_outcomes()) %>%
  # Group infrequent levels of categorical predictors into "other"
  step_other(all_nominal(), -all_outcomes()) %>%
  # Convert categorical predictors into dummy variables
  step_dummy(all_nominal(), -all_outcomes()) %>%
  # Remove near-zero variance predictors
  step_nzv() %>%
  # Upsample minority class to address class imbalance
  themis::step_upsample(y, over_ratio = 0.9, skip = TRUE, seed = 123) %>%
  # Subsample dataset to reduce size
  step_sample(size = 5000, skip = TRUE, seed = 456)

```
We can ignore the warning since the selectors not being used wont be a problem for the project

## Resampling

Set 5 groups (folds) with the same 'default' variable distribution

```{r}
set.seed(1234)
cv_data <- vfold_cv(train, v = 5, repeats = 1, strata = default)
cv_data
```

## Decision Tree

```{r}
# Create the Decision Tree model
tree_model <- rpart(default ~ ., data = Datos_MOD1)
```


## Grid Search

```{r}
# Define the hyperparameter grid search
grid <- expand.grid(cp = seq(0.01, 0.1, by = 0.01))

# Perform hyperparameter search using cross-validation
tree_model <- train(default ~ ., data = Datos_MOD1, method = "rpart", tuneGrid = grid, trControl = trainControl(method = "cv"))

# Visualize the results of hyperparameter search
print(tree_model)
```


## Training with Optimal Value

```{r}
optimal_cp <- 0.1
tree_model <- rpart(default ~ ., data = train, control = rpart.control(cp = optimal_cp))

predictions <- predict(tree_model, newdata = test, type = "class")

# Evaluate the model on the 'test' set
confusion_matrix <- table(predictions, test$default)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)

# Visualize the confusion matrix
print(confusion_matrix)

# Show the model accuracy
print(paste("Accuracy:", accuracy))

```

## Calculate sensitivity

```{r}
sensitivity <- confusion_matrix[2, 2] / sum(confusion_matrix[2, ])
specificity <- confusion_matrix[1, 1] / sum(confusion_matrix[1, ])
print(paste("Sensitivity:", sensitivity))
print(paste("Specificity:", specificity))
```

## Conclusions

### Data Exploration:
- We observed that the target variable "default" is imbalanced, with approximately 22.12% of clients expected to default next month.
- The dataset contains information about demographic factors such as gender, education, and marital status, as well as payment history and credit amounts.

### Data Preprocessing:
- We converted categorical variables into factors and normalized numeric predictors.
- Handling imbalanced data, we upsampled the minority class and subsampled the dataset to reduce its size.

### Model Training and Evaluation:
- We split the dataset into training and testing sets with an 80/20 ratio.
- Utilizing cross-validation with 5 folds and grid search, we tuned hyperparameters for a decision tree model.
- The model achieved an accuracy of approximately 73.6% on the test set.
- Sensitivity, which measures the proportion of actual positives that are correctly identified as such, was found to be 28.3%, indicating the model's ability to detect true defaults.
- Specificity, representing the proportion of actual negatives that are correctly identified, was calculated at 85.5%, suggesting the model's capacity to recognize true non-defaults.

Overall, the decision tree model demonstrates moderate performance in predicting credit card defaults. Further optimization and exploration of alternative algorithms may enhance model accuracy and predictive power.




